---
title: "Repeatability of flowering phenology in *Echinacea angustifolia*"
output: html_document
bibliography: echinacea_repeatability.bib
---

## Introduction

Previous research has suggested that flowering earlier is selectively favored, both in *E. angustifolia* and across other plant species [@Munguia-Rosas2011]. If long-lived polycarpic perennials have consistent flowering phenologies, we could assume those individuals have consistent

## Methods

Plants that have only flowered once during the study period were excluded from the analysis (XX individuals). 

Since year-to-year fluctuations in weather, as well as spring burning, can affect overall timing of flowering phenology (figure X), year was included as a fixed effect in the model. 

## Results

Repeatibility of flowering time (start date):

```{r}
library(rptR)
load("data/phen_dataset.rda")
r1 <- rpt(startNum ~ year + (1|cgPlaId), grname = "cgPlaId", data = p1996, datatype = "Gaussian",
                  nboot = 1000)
summary(r1)
plot(r1)
```

Repeatability of flowering duration:

```{r}
r2 <- rpt(dur ~ year + (1|cgPlaId), grname = "cgPlaId", data = p1996, datatype = "Gaussian",
                  nboot = 1000)
summary(r2)
plot(r2)
```

## Discussion


## References